\documentclass[a4paper]{extarticle}
\usepackage{amsfonts} % cursed Z symbol
\usepackage{extsizes}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{float}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{cancel}
\usepackage[table,xcdraw]{xcolor}
\usepackage[a4paper, margin=0.5in]{geometry}
\usepackage{paracol}
% :map = :s#.*#\\pushnote{&}{() ``''}^Mj
% :se guifont=AR\ PL\ KaitiM\ Big5\ Bold\ 20


\usepackage{xeCJK}
\setCJKmainfont{AR PL UMing CN}


\newcounter{notecounter}
\newwrite\savefile
\immediate\openout\savefile=tempfile.tmp
\newcommand{\pushnote}[2]{
\unskip% removes any preceding space
\stepcounter{notecounter}%
\colorbox{gray!30}{{\Large #1}}%
\raisebox{2ex}{\makebox[2pt]{\tiny\thenotecounter}}%
\immediate\write\savefile{\thenotecounter : \space #1 \space #2

}%
\ignorespaces% removes any following space
}

\newcommand{\bothl}[2]{
{\large #1}
\switchcolumn
#2
\switchcolumn*
}

\newcommand{\outputnotes}{ \immediate\closeout\savefile \vspace{0.5cm} \large \input{tempfile.tmp} \vspace{0.5cm}
\immediate\openout\savefile=tempfile.tmp
\setcounter{notecounter}{0}
}

\newcommand{\separator}{
\end{paracol}

\outputnotes

\begin{paracol}{2}
}


\begin{document}

\section{Beginner}

\subsection{To Live}
\vspace{1cm}
\begin{paracol}{2}
\bothl{活着}{To Live}
\bothl{余华}{Yu Hua}
\vspace{1cm}
\end{paracol}


\begin{paracol}{2}
\bothl{\pushnote{我}{(wǒ) ``I; me; my''}\pushnote{比}{(bǐ) ``to compare; (followed by a noun and adjective) more `adj.' than `noun'''}\pushnote{现在}{(xiàn zài) ``now; at present; at the moment; modern; current; nowadays''}\pushnote{年轻}{(nián qīng) ``young''}\pushnote{十岁}{(shí suì) ``ten'' + ``age''=``ten years of age''}\pushnote{的}{(de) possessive particle}\pushnote{时候}{(shí hou) ``time; length of time; moment; period''}，}{When I was ten years younger than I am now}

\separator

\bothl{\pushnote{获得}{(huò dé) ``to obtain; to receive; to get''}\pushnote{了}{(le) completed action marker}\pushnote{一个}{(yī) ``one'' + (gè), classifier}\pushnote{游手好闲}{(yóu shǒu hào xián) idiom: ``to idle about''; in this case 好 is (hào) ``to be fond of; to have a tendency to; to be prone to''}的\pushnote{职业}{(zhí yè) ``occupation; profession; vocation; professional''}，

去\pushnote{乡间}{(xiāng jiān) ``in the country; rural; pastoral''}\pushnote{收集}{(shōu jí) ``to gather; to collect''}\pushnote{民间}{(mín jiān) ``among the people; popular; folk''}\pushnote{歌谣}{(gē yáo) ``folksong/ballad/nursery rhyme''}。}{I had the carefree job of going to the countryside to collect popular folk songs.}

\separator

\bothl{那一年的\pushnote{整个}{(zhěng gè) ``whole; entire; total''}夏天，我\pushnote{如同}{(rú tóng) ``like; as''}一只\pushnote{乱}{(luàn) ``in confusion or disorder; in a confused state of mind; disorder; upheaval; riot''}飞的\pushnote{麻雀}{(má què) ``sparrow; mahjong''}，}{That year, for the entire summer, I was like a sparrow soaring recklessly.}

\separator

\bothl{\pushnote{游荡}{(yóu dàng) ``to wander; to roam about; to loaf about; to be idle''}在\pushnote{知了}{(zhī liǎo) ``cicada''}和\pushnote{阳光}{(yáng guāng) ``sunshine''}\pushnote{充斥}{(chōng chì) ``to be full of; to flood; to congest''}的\pushnote{村舍}{(cūn shè) ``cottage''}

\pushnote{田野}{(tián yě) ``field; open land''}。}{I would wander amid the village houses and the open country, which was full of cicadas and flooded with sunlight.}

\separator

\bothl{我\pushnote{喜欢}{(xǐ huan) ``to like; to be fond of''}喝\pushnote{农民}{(nóng mín) ``peasant; farmer''}那种\pushnote{带有}{(dài yǒu) ``to have as a feature or characteristic; to have an element of (confidence, sweetness, malevolence etc); to carry''}\pushnote{苦味}{(kǔ wèi) ``bitter taste; bitterness''}的茶水，}{I had a special affection for that bitter tea that farmers brew. }

\separator


\bothl{他们的茶\pushnote{桶}{(tǒng) ``bucket; (trash) can; barrel (of oil etc)''}\pushnote{田埂}{(tián gěng) ``embankment between fields''}的树下，我\pushnote{毫无}{(háo wú) ``not in the least; to completely lack''}\pushnote{顾忌}{(gù jī) ``to have misgivings; apprehension; worry; qualm; scruple''}地\pushnote{拿起}{(ná qǐ) ``to pick up''}\pushnote{漆}{(qī) ``paint; lacquer; to paint (furniture, walls etc)''}\pushnote{满}{(mǎn) ``to fill; full; filled; packed; fully; completely; quite; to reach the limit; to satisfy; satisfied; contented''}\pushnote{茶垢}{(chá gòu) ``tea stain (on the inside of a tea pot, tea cup etc)''}的\pushnote{茶碗}{(chá wǎn) ``teacup''}\pushnote{舀水}{(yǎo shǔi) ``to ladle water; to scoop up water''}喝，}{There would always be a bucket of just that kind of tea under a tree by the ridge between the fields, and without a second thought I would ladle out enough to fill my tea-stained bow.}

\separator

\bothl{还把\pushnote{自己}{(zì jǐ) ``oneself; one's own''}的\pushnote{水壶}{(shǔi hú) ``kettle; canteen; watering can''}\pushnote{灌}{(guàn) ``to irrigate; to pour''}满，与田里\pushnote{干活}{(gàn huǒ) ``to work; to be employed''}的\pushnote{男人}{(nán rén) ``man; male''}说上几句\pushnote{废话}{(fèi huà) ``nonsense; rubbish; superfluous words''}，}{Once I'd filled it to the brim, I'd start bullshitting with some of the male workers.}


\separator


\bothl{在\pushnote{姑娘}{gū niang() ``girl; young woman; young lady; daughter''}因我而起的\pushnote{窃窃}{(qiè qiè) ``privately; secretly; unobtrusively''}\pushnote{私}{(sī) ``personal; private; selfish''}笑里\pushnote{扬长而去}{(yáng cháng ér qù) idiom: ``to swagger off; to walk off (or drive off etc) without a second thought for those left behind''}。}{The girls would whisper among themselves and then stifle their chuckles as I'd swagger of.}


\separator

\bothl{我\pushnote{曾经}{(céng jīng) ``once; already; former; previously; ever; (past tense marker used before verb or clause)''}和一位\pushnote{守}{(shǒu) ``to guard; to defend; to keep watch; to abide by the law; to observe (rules or ritual); nearby; adjoining''}着瓜田的老人\pushnote{聊}{(liáo) ``(coll.) to chat; (literary) temporarily; for a while; (literary) somewhat; slightly; (literary) to depend upon''}了\pushnote{整整}{(zhěng zhěng) ``/whole; full''}一个\pushnote{下午}{(xià wǔ) ``afternoon''}，}{I once spent a whole aftemoon talking with an old man who kept a melon patch.}

\separator

\bothl{这是我\pushnote{有生以来}{(yǒu shēng yǐ lái) ``since birth; for one's whole life''}瓜吃得\pushnote{最多}{(zuì duō) ``at most; maximum; greatest (amount); maximal''}的一次，}{I ate more melons that day than I ever had in my life.}
\bothl{当我\pushnote{站起来}{(zhàn qǐ lai) ``to stand up''}\pushnote{告辞}{(gào cí) ``to say goodbye; to take one's leave''}时，}{When I stood up to leave, }

\separator

\bothl{\pushnote{突然}{(tū rán) ``sudden; abrupt; unexpected''}\pushnote{发现}{(fā xiàn) ``to notice; to become aware of; to discover; to find; to detect; a discovery''}自己像个\pushnote{孕妇}{(yùn fù) ``pregnant woman''}\pushnote{一样}{(yī yàng) ``same; like; equal to; the same as; just like''}

\pushnote{步履}{(bù lǚ) ``gait; to walk''}\pushnote{艰难}{(jiān nán) ``difficult; hard; challenging''}了。}{I suddenly realized that I had as much difficulty walking as a pregnant woman.}

\separator

\bothl{\pushnote{然后}{(rán hòu) ``then; after that; afterwards''}我与一位\pushnote{当上}{(dāng shang) ``to take up duty as; to assume a position; to assume; to take on (an office)''}了\pushnote{祖母}{(zǔ mǔ) ``father's mother; paternal grandmother''}的\pushnote{女人}{(nǚ rén) ``woman''}坐在\pushnote{门槛}{(mén kǎn) ``doorstep; sill; threshold''}上，}{Later that day, I sat on the porch with a woman who had already become a grandmother.}

\bothl{她\pushnote{编}{(biān) ``to weave; to plait; to organize; to group; to arrange; to edit; to compile; to write; to compose; to fabricate; to make up''}着\pushnote{草鞋}{(cǎo xié) ``straw sandals''}为我\pushnote{唱}{(chàng) ``to sing; to call loudly; to chant''}了一支《十月\pushnote{怀胎}{(huái tāi) ``to become pregnant; to carry a child in the womb''}》。}{As she weaved a pair of straw sandals she sang “Ten Month Pregnancy” for me.}

\separator

\bothl{我最喜欢的是\pushnote{傍晚}{(bàng wǎn) ``dusk''}来到时，坐在农民的\pushnote{屋}{(wū) ``house, room''}前，}{What I loved most was sitting before the peasants’ houses just as dusk fell.}

\bothl{看着他们将提上的井水\pushnote{泼}{(pō) ``to splash; to spill; rough and coarse; brutish''}在地上，\pushnote{压住}{(yā zhù) ``to press down; to crush down; to restrain (anger); to keep down (voice)''}

\pushnote{蒸腾}{(zhēng tēng) ``(of a vapor etc) to rise; to hang in the air''}的\pushnote{尘土}{(chén tǔ) ``dust''}，\pushnote{夕阳}{(xī yáng) ``sunset; the setting sun''}的\pushnote{光芒}{(guāng máng) ``rays of light; brilliant rays; radiance''}在\pushnote{树梢}{(shù shāo) ``the tip of a tree; treetop''}上

\pushnote{照射}{(zhào shè) ``to shine on; to light up; to irradiate''}下来，}{As the sun's rays came down through the delicate branches, I would watch the peasants pour well water onto the ground, cooling the hot dust and sand.}

\separator

\bothl{拿一把他们递过来的\pushnote{扇子}{(shàn zi) ``fan''}，尝尝他们和盐一样咸的\pushnote{咸菜}{(xián cài) ``salted vegetables; pickles''}
，}{Holding the fan they passed over to me, I would try the pickled vegetables, which always tasted like salt.}
\bothl{看看几个年轻女人，和男人们说着话。}{I would watch the girls and talk with the men.}
\end{paracol}

\outputnotes



\section{Intermediate}
\subsection{2001: A Space Odyssey}
\vspace{1cm}
By Sir Arthur Charles Clarke, published in 1968.
\vspace{1cm}


\begin{paracol}{2}
\bothl{2001: A Space Odyssey}{
\pushnote{太空}{(tài kōng) ``the firmament; outer space; aether; firmament''}
\pushnote{漫游}{(màn yóu) ``roam; cruise; go on a pleasure trip''}
}

\separator


\bothl{太初之夜 \vspace{0.5cm} }{PRIMEVAL NIGHT}
\bothl{1 灭绝之路 \vspace{0.5cm} }{1 – The Road to Extinction}


\bothl{这时，\pushnote{干旱}{(gān hàn) ``drought; arid; dry; aridity''}
已经\pushnote{持续}{(chíxù) ``to continue; to persist; to last; sustainable; preservation''}
了一千万年，可怕的恐龙也早已结束了\pushnote{主宰}{(zhǔ zǎi) ``dominate''; 宰 zǎi is ``to slaughter; to butcher; to kill (animals etc)/(coll.)'' as in 屠宰場 (tú zǎi chǎng) ``slaughterhouse'' }。}{The drought had lasted now for ten million years, and the reign of the terrible lizards had long since ended.}
\bothl{在\pushnote{赤道}{(chì dào) ``Equator''; 赤 (chì) is  ``red, scarlet, bare, naked'' as in 赤脚 (chì jiǎo) ``bare feet''}此处，
日后将以非洲之名而\pushnote{闻名}{(wén míng) ``famous''}的这块大陆上，
求生之战的\pushnote{凶残}{(xiōng cán) ``fierce, brutal''}，
已\pushnote{沸腾}{(fèi téng) ``boiling, to boil''}到新的高点，
胜出者则尚未见\pushnote{踪影}{(zōng yǐng) ``trace, vestige''; 影 is ``shadow'', see 电影 (diàn yǐng) ``movie''}。
}{Here on the Equator, in the continent which would one day be known as Africa, the battle for existence had reached a new climax of ferocity, and the victor was not yet in sight.
}

\separator

\bothl{在这片\pushnote{干枯}{(gān kū) ``Dry and withered''}
的\pushnote{不毛之地}{(bù máo zhī dì) idiom: `` A barren and barren place without crops, abandoned land '' }
上，想要\pushnote{繁衍}{(fán yǎn) ``reproduction''}
下去，或者起码有点\pushnote{存活}{(cún huó) ``survive''}
下去的\pushnote{指望}{(zhǐ wang) ``hope'' }
，就得要小，要快，要狠。\pushnote{狠}{hěn) ``ruthless; fierce; ferocious; determined; resolute; to harden (one's heart)''}
}{In this barren and desiccated land, only the small or the swift or the fierce could flourish, or even hope to survive.}
\bothl{\pushnote{荒野}{(huāng yě) ``wilderness'' }
上的\pushnote{猿人}{(yuán rén) ``Ape man'' }
够不上这些条件，所以没的繁衍。
}{The man-apes of the veldt were none of these things, and they were not flourishing.}

\separator

\bothl{再说明白点，他们已经离\pushnote{灭种}{(miè zhǒnɡ) ``extermination, extinction'' }
不远。}{Indeed, they were already far down the road to racial extinction.}
\bothl{他们五十来个，\pushnote{盘踞}{(pán jù) ``occupy'' } 
了一些山洞。

\pushnote{俯视}{(fǔ shì) ``Looking down'' from 俯 ``to look down; bow down'' and 视 ``look ; vision''} 
而下，是一个干枯的小\pushnote{盆地}{(pén dì) ``basin'' } 
。
盆地里流过一条\pushnote{迟滞}{(chí zhì) ``sluggish; delay'' } 
的小溪，是来自北方两百英里外山上的融雪。
}{About fifty of them occupied a group of caves overlooking a small,
parched valley, which was divided by a sluggish stream fed from snows
in the mountains two hundred miles to the north.
}
\bothl{干旱厉害的时候，小溪\pushnote{彻底}{(chè dǐ) ``thorough; thoroughgoing; downright; every inch''}\pushnote{蒸发}{(zhēng fā) ``evaporate; evaporation; ablate; exhalation'' }
，这个\pushnote{部落}{(bù luò) ``tribe''}
就活在焦渴的阴影里。
}{In bad times the stream vanished completely, 
and the tribe lived in the shadow of thirst.
}

\separator

\bothl{他们本来就\pushnote{饿惯}{饿 (è) ``hunger'', 惯 (guàn) ``be used to; be in the habit of; indulge'': ``Be used to hunger''}
了，现在则\pushnote{濒临}{(bīn lín) ``be close to; border on; be on the verge of''}
饿死。}
{It was always hungry, and now it was starving.}
\bothl{当黎明第一道\pushnote{朦胧}{(méng lóng) ``Obscure'' }\pushnote{曙光}{(shǔ guāng) ``Dawn'' }\pushnote{掩入}{(yǎn rù) ``Cover in'' }
山洞的时候，\pushnote{望月者}{ from 望 (wàng) ``Towards, at'' + 月 (yuè) ``Moon'' + 者 (zhě) ``person'' = "He who looks at the Moon", "Moon-watcher"}
发现父亲已经在夜里死了。
}{When the first faint glow of dawn crept into the cave,
Moon-Watcher saw that his father had died in the night. }
\bothl{他并不明白``这个老东西''就是他的父亲，父子关系还不在他的理解范围之内。}
{He did not know that the Old One was his father, for such a
relationship was utterly beyond his understanding, }
\bothl{\pushnote{然而}{(rán ér) ``However, but'' }
当他看到那具\pushnote{羸弱}{(léi ruò) ``Weak and frail'' }
的尸体时，心里还是\pushnote{隐约}{(yǐn yuē) ``Faint; indistinctint; vaguely'' }
感到一阵不安——后来，这种不安才会演化为\pushnote{哀伤}{(āi shāng) ``sad; distressed; grieved; feel sorrow or grief'' }
。
}{but as he looked at the emaciated body he felt dim disquiet
that was the ancestor of sadness.}

\separator

\bothl{两个孩子饿得一直低声\pushnote{哭泣}{( kū qì) ``cry; weep; sob; blubber'' }
，望月者\pushnote{吼}{(hǒu) ``Roar; howl'' }
了一声\pushnote{止住}{(zhǐ zhù) ``stop; halt; desist; bring to a stop;'' }
他们。
}{The two babies were already whimpering for food, but became silent
when Moon-Watcher snarled at them. }
其中一个孩子的妈妈，为了护她没法好好\pushnote{喂养}{(wèi yǎng) ``feed; raise; keep'' }
的\pushnote{婴孩}{(yīng hái) ``baby; child'' }
，\pushnote{愤怒}{(fèn nù) ``anger; rage; wrath; indignation; bile'' }
\bothl{地朝他回吼了一声。}{One of the mothers, defending the
infant she could not properly feed, gave him an angry growl in return; 
}
\bothl{但他连\pushnote{揍}{(zòu) ``beat; break; smash'' }
她一\pushnote{拳}{(quán) ``boxing; fist; pugilism'' }
、修理她\pushnote{放肆}{(fàng sì) ``unbridled; wanton; impudicity; presumption'' }
的力气都没有了。
}{he lacked the energy even to cuff her for her presumption.}
\bothl{现在天色亮得可以出发了。}{Now it was light enough to leave.}

\separator

\bothl{望月者\pushnote{拖}{(tuō) ``drag; pull; mop; delay ; draw'' }
着那具\pushnote{干枯}{(gān kū) ``withered; dried-up; wizened; shrivelled'' }
的尸体，

\pushnote{弯腰}{(wān yāo) ``stoop; bow; stoop down'' }
钻出头顶有片\pushnote{斜}{(xié) `` inclined ; skew; slanting; bevel'' }\pushnote{岩}{(yán) ``rock; cliff; crag'' }\pushnote{延伸}{(yán shēn) ``extend; stretch; elongate; rolling out'' }
出去的洞口。
}{Moon-Watcher picked up the shriveled corpse and dragged it after him 
as he bent under the low overhang of the cave.}
\bothl{出了山洞，他把尸体
\pushnote{扛}{(káng) ``shoulder; carry on the shoulder''}
在\pushnote{肩上}{() ``''}
，站直了身体——在这个世界上，
还只有他这种动物有这个\pushnote{本领}{() ``''}
。}{Once outside, he threw the body over his 
shoulder and stood upright – the only animal in all this world able 
to do so.}
\bothl{比起他的同类，望月者几乎算是个巨人。}
{Among his kind, Moon-Watcher was almost a giant. }

\separator

\bothl{近五\pushnote{英尺}{(yīng chǐ) ``foot (pl. feet)'', from 
英 (yīng) ``England'' and 尺 (chǐ), an old
Chinese unit of measure of almost one third of a meter}
高，尽管营养不良，还有一百多\pushnote{磅重}{磅 (bàng) ``cut'', an old character here used for onomatopoeia + 重 (zhòng) ``weight''=磅重 ``English pound''}
。}{He was nearly five feet high, and though badly undernourished weighed over a hundred pounds.}
\bothl{他\pushnote{毛茸茸}{(máo rōng rōng) ``hairy; brushy; downy; flocky''}
的身体，肌肉发达，介于人与猿之间，但他的头，则近乎人而非猿；
}{His hairy, muscular body was halfway between ape and man,
but his head was already much nearer to man than ape.}
\bothl{额头很低，\pushnote{眼窝}{(yǎn wō) ``eye socket''}\pushnote{深}{(shēn) ``deep; difficult; profound; thoroughgoing''}\pushnote{陷}{(xiàn) ``trap; pitfall; defect; deficiency; sink''}
，不过，他的\pushnote{基因}{(jī yīn) ``gene; genome''}
里无疑已具备演化为人类的希望。
}{The forehead was low, and there were ridges over the eye sockets,
yet he unmistakably held in his genes the promise of humanity.}
\bothl{
当他望着\pushnote{更新世}{(gēng xīn shì) ``Pleistocene (geological epoch from 2m years ago, covering the most recent ice ages)''}
这个\pushnote{残酷}{(cán kù) ``cruel; brutal; ruthless; inhuman; bloodiness''}
的世界时，

\pushnote{眼神}{(yǎn shén) ``expression in one's eyes; light; eyesight''}
已经远非猿类可及。
}{As he looked out upon the hostile world of the Pleistocene, there was already something in his gaze beyond the capacity of any ape.}

\separator

\bothl{在他\pushnote{黝黑}{(yǒu hēi) ``dark; swarthy''}\pushnote{深邃}{(shēn suì) ``profound; abstruse; deep; recondite''}
的双眼里，\pushnote{透}{(tòu) ``penetrate; pass [seep] through; tell secretly; ''}
着一种\pushnote{逐渐}{(zhú jiàn) ``gradually; by degrees; little by little''}\pushnote{苏醒}{(sū xǐng) ``revive; regain consciousness''}
的\pushnote{知觉}{(zhī jué) ``consciousness; perception''}
——一种不经多代演化\pushnote{不足}{(bù zú) ``insufficient; lacking; deficiency; not enough; inadequate; not worth; cannot; should not''}
以具现、要灭绝则快得很的智能，在其中有了最初的\pushnote{闪烁}{(shǎn shuò) ``flicker; twinkle; evasive; vague''}
。}{In those dark, deep-set eyes was a dawning awareness – the first intimations of an intelligence that could not possibly fulfill itself for ages yet, and might soon be extinguished forever.}

\separator

\bothl{四下没有危险的\pushnote{迹象}{(jì xiàng) ``sign; evidence; indication; token; hide''}
，于是望月者\pushnote{沿}{(yán) `` along, follow; trim''}
着洞外\pushnote{近乎}{(jìn hu) ``almost''}\pushnote{垂直}{(chuí zhí) ``vertical; perpendicular; squareness; verticality;''}
的\pushnote{陡坡}{(dǒu pō) ``steep slope; heavy gradient; abrupt slope; bank'';
陡 is  ``steep; precipitous; abrubtly; suddenly; unexpectedly'', see 
陡然 (dǒu rán) ``suddenly; unexpectedly; abruptly; precipitously; stumbling''
}爬下，身上背的尸体没有\pushnote{造成}{(zào chéng) ``cause; give rise to; bring about; create; work''}
太大\pushnote{妨碍}{(fáng ài) ``hinder; impede; hamper; obstruct; bar''; 
see also 障礙 (zhàng ài) ``barrier; obstruction; hindrance; impediment; obstacle''}
。}{There was no sign of danger, 
so Moon-Watcher began to scramble down the almost vertical 
slope outside the cave, only slightly hindered by his burden.}

\separator

\bothl{
\pushnote{部落}{(bù luò) ``tribe; clan''}
里其他的猿人，似乎一直在等待他的信号般，\pushnote{纷纷}{(fēn fēn) ``one after another; in succession''}
从\pushnote{岩壁}{(yán bì) ``a wall of rock''}
下方自己的\pushnote{洞穴}{(dòng xué) ``cave; cavern; burrow; grotto''}
里\pushnote{钻出来}{(zuàn chū lái) ``come out; emerge''}
，\pushnote{急急忙忙}{(jí jí máng máng) ``hurried and busy; in a rush; in great hast''}
赶向那条\pushnote{泥泞}{(ní nìng) ``muddy; slushy; miry; sloppy; loblolly''}
的小溪

，\pushnote{寻觅}{(xún mì) ``seek; look for''}
他们早上要喝的水。}{As if they had been waiting for his signal,
the rest of the tribe emerged from their own homes farther down 
the rock face, and began to hasten toward the muddy waters of 
the stream for their morning drink.}

\separator

\bothl{望月者望过谷地，看看是否有“对方”出现。}{Moon-Watcher 
looked across the valley to see if the Others were in sight,}
\bothl{但没有踪影。}{but there was no trace of them.}
\bothl{也许他们还没有离开自己的洞穴，也许已经沿着山腰去他处
\pushnote{觅食}{(mì shí) ``foraging''; this is the same 觅 seen above in 寻觅}
了。} {Perhaps they had not yet left their caves, or were already 
foraging farther along the hillside. }


\bothl{\pushnote{既然}{(jì rán) ``since; as; now that; in as much''}
不见踪影，望月者就把他们忘在脑后。}{Since they were nowhere to be seen, Moon-Watcher forgot them;}


\bothl{他还没有能力同时
\pushnote{操心}{(cāo xīn) ``bother ; worry about; trouble about; take pains''}
一件以上的事情。}{he was incapable of worrying about more than one thing at a time.}

\separator

\bothl{首先他得解决这个老东西，但这个问题不用花什么
\pushnote{脑筋}{(nǎo jīn)  ``brains; mind; head; way of thinking''}
。}
{First he must get rid of the Old One, but this was a problem 
that demanded little thought. }
\bothl{这一季里，死的\pushnote{同伴}{(tóng bàn) ``companion; comrade; fellow''}
很多。}{There had been many deaths this
season, }
\bothl{之前，他自己洞里就已经死了一个。}{, one of them in his own cave; }
\bothl{他只要在上次
\pushnote{弦月}{(xián yuè) ``half-moon; the 7th and 8th and 22nd and 23rd of the lunar month''}
\pushnote{时分}{(shí fēn) ``time; period during the day''}
扔下那个新生婴儿的地方放下这具\pushnote{尸首}{(shī shou)  ``corpse; carcass; dead body''}
，\pushnote{鬣狗}{(liè gǒu) ``hyena''}
就会解决剩余问题。}{he had only to put the corpse where he had left the new baby at the last quarter of the moon, and the hyenas would do the rest.
}
\bothl{鬣狗好像知道他要来，已经在这小山谷和
\pushnote{疏}{(shū) ``dredge ; neglect; disperse; scatter; sparse''}
林草原的交口上等着了。}
{They were already waiting, where the little valley fanned out into the savanna, almost as if they had known that he was coming.}
\bothl{望月者把尸体丢在一棵
\pushnote{灌木}{(guàn mù) ``bush; shrub''}
下——先前的骨头都已经不见——然后就急急赶回部落。}{Moon-Watcher left the body under a small bush – all the earlier bones were already gone – and hurried back to rejoin the tribe.}
\bothl{从此，望月者再没有想起过他的父亲。} {He never thought of his father again.}

\separator

\bothl{他的两个\pushnote{配偶}{(pèi ǒu) ``spouse''}
、其他洞穴出来的成年同类，以及大多数的少年同类，正沿山谷而上，在那些被干旱\pushnote{摧残}{(cuī cán) ``to ravage; to ruin''}
的林木间觅食，找一些浆果、多汁的树根和树叶，以及偶尔意外\pushnote{捕获}{(bǔ huò) ``to catch; to capture; to seize''}
的小\pushnote{蜥蜴}{(xī yì) ``lizard''}
和\pushnote{啮齿动物}{(niè chǐ dòng wù) ``rodent''}
。}{His two mates, the adults from the other caves, and most of the youngsters were foraging among the drought-stunted trees farther up the valley, looking for berries, succulent roots and leaves, and occasional windfalls like small lizards or rodents. }
\bothl{只有小婴儿和\pushnote{虚弱}{(xū ruò) ``weak; in poor health''}\pushnote{不堪}{(bù kān) ``cannot bear; cannot stand; utterly; extremely''}的老家伙才留在洞穴里。
觅食一天之后如果还有剩余，也许还可以喂他们吃一点。
如果没有，鬣狗则很快又要走\pushnote{走运}{(zǒu yùn) ``to have good luck; to be in luck''}了。}{Only the babies and the feeblest of the old folk were left in the caves; if there was any surplus food at the end of the day's searching, they might be fed. If not, the hyenas would soon be in luck once more.}
\bothl{不过今天是很棒的一天——虽然望月者对过去并没有什么真正的记忆，也没法把这一次和其他时候相比较。}
{But this day was a good one – though as Moon-Watcher had no real remembrance of the past, he could not compare one time with another. }

\separator

\bothl{他在一棵枯树根上发现了一个\pushnote{蜂窝}{(fēng wō)  ``honeycomb; beehive''}，因而享受了一顿他们\pushnote{族类}{(zú lèi) ``clan; race''}\pushnote{前所未知}{(qián suǒ wèi zh1) idiom: ``new experience, previously unknown''}的\pushnote{无上美味}{(wǔ shàng měi wèi) idiom: ``the ultimate delicacy''}。
傍晚时分，他带着大伙回家的时候，还不时\pushnote{舔舔}{(tiǎn tiǎn) ``to lick; to lap''}手指。}{He had found a hive of bees in the stump of a dead tree, and so had enjoyed the finest delicacy that his people could ever know; he still licked his fingers from time to time as he led the group homeward in the late afternoon.}
\bothl{当然，他也被\pushnote{蜇}{(zhē) ``to sting''}了好几下，但他没有太在意。}{Of course, he had also collected a fair number of stings, but he had scarcely noticed them.}
\bothl{现在他几乎可以说从没这么心满意足过，因为虽然还是饿，但已经不会饿得虚软。}{He was now as near to contentment as he was ever likely to be; for though he was still hungry, he was not actually weak with hunger.}
\bothl{对猿人来说，\pushnote{夫复何求}{(fū fù hé qiú) idiom: to not need enything anymore``''}。}{That was the most to which any man-ape could ever aspire.}
\end{paracol}

\outputnotes

\subsection{Moby Dick (白鲸)}


\vspace{1cm}
by Herman Melville, published in 1851

\vspace{1cm}
from Chapter 16: The Ship.

\vspace{1cm}

\begin{paracol}{2}
\bothl{“您就是‘\pushnote{裴廓德号}{Transliteration: 裴 (péi), a surname; ``(of a garment) long and flowing'' + 
廓 (kuò) ``(bound form) extensive; vast; (bound form) outline; general shape; (bound form) to expand; to extend; '' +
德 (dé) ``virtue; goodness; morality; ethics; kindness; favor; character; kind'' +
号 (hào) ``ordinal number; day of a month; mark; sign; business establishment; size; ship suffix (in this case in particular); horn (wind instrument); bugle call; assumed name; to take a pulse; classifier used to indicate number of people''.}’的船长吗？”我走到\pushnote{帐篷}{(zhàng peng) ``tent''}门口问了一声。}{“Is this the Captain of the Pequod?” said I, advancing to the door of the tent.}
\bothl{
“就算我是，你找船长干什么？”他\pushnote{咄咄逼人}{(duō duō bī rén) ``overbearing; forceful; aggressive; menacing; imperious''}。
}{
“Supposing it be the captain of the Pequod, what dost thou want of him?” he demanded.
}
\bothl{
“我想当水手。”
}{
“I was thinking of shipping.”
}

\bothl{
“你想当，是吗？我看你不像\pushnote{南塔基}{Transliteration: 南 (nán) ``South'' +
塔 (tǎ) ``pagoda'' + 基 (j1) ``base; foundation''}人——你在\pushnote{漏水}{(lòu shuǐ) ``to leak (of water)''}的船上干过水手？”
}{
“Thou wast, wast thou? I see thou art no Nantucketer—ever been in a stove boat?”
}

\bothl{
“没有，先生，从没干过。”
}{
“No, Sir, I never have.”
}

\bothl{
“对\pushnote{捕鲸}{(bǔ jīng) ``whaling'', from
捕 ``to catch; to seize; to capture'' and 
鲸 ``whale''.  N.B. the word for ``whale'' is 鲸鱼 (jīng yú)}这行当\pushnote{一窍不通}{(yī qiào bù tōng) ``lit. doesn't (even) enter a single aperture (of one's head); I don't understand a word (idiom); it's all Greek to me''}，我敢肯定——\pushnote{呃}{(è) ``exclamation ; hiccup''}？”
}{
“Dost know nothing at all about whaling, I dare say—eh?
}
\end{paracol}

\outputnotes

\begin{paracol}{2}
\bothl{“是的，先生。但我保证，我很快就可以学会。我在\pushnote{商船}{(shāng chuáng) ``merchant ship''}上当过几回水手，我想——”}{“Nothing, Sir; but I have no doubt I shall soon learn. I've been several voyages in the merchant service, and I think that—”}
\bothl{“该死的商船水手。}{“Merchant service be damned.}
\bothl{别跟我说那些\pushnote{鬼话}{(gǔi huà) ``lie; false words; nonsense''}。}{ Talk not that lingo to me.}
\bothl{你看那条腿？}{ Dost see that leg?—}
\bothl{你再敢跟我\pushnote{胡扯}{(hú chě) ``to chatter; nonsense; blather''; 扯 is ``to pull; to tear; (of cloth, thread etc) to buy; to chat; to gossip (coll.)''}商船水手的事，我就叫你的腿和屁股分家\pushnote{啰}{(luo) ``(final exclamatory particle)'', it's the 啰 in 哈啰 (hā luō) ``hello''}。}{I'll take that leg away from thy stern, if ever thou talkest of the marchant service to me again. }
\bothl{商船水手，算什么！}{Marchant service indeed! }
\bothl{我看你现在对干过商船水手很\pushnote{自豪}{(zì háo) ``proud (of one's achievements etc)''}吧。}{I suppose now ye feel considerable proud of having served in those marchant ships.}
\bothl{不过算你走运，\pushnote{伙计}{(huǒ jī) ``partner; fellow; mate; waiter; servant; shop assistant''}！你是为什么想要去捕鲸的呢，呃？}{ But flukes! man, what makes thee want to go a whaling, eh?—}
\bothl{这很可疑，不是吗？你说。}{it looks a little suspicious, don't it, eh?—}
\bothl{——你干过海盗，是吗？——}{Hast not been a pirate, hast thou?—}
\bothl{你\pushnote{抢劫}{(qiǎng jié) ``to rob; looting'', see also 劫机 (jié jī) ``hijacking; air piracy''}过那商船的船长，有这事吗？——}{Didst not rob thy last Captain, didst thou?—}
\bothl{你出海的时候，会对船上的\pushnote{头头}{(tóu tóu) ``head; chief''}们起\pushnote{歹}{(dǎi) ``bad; wicked; evil''}念吗？”}{Dost not think of murdering the officers when thou gettest to sea?”}

\separator


\bothl
{我表情\pushnote{坚定}{(jiān dìng) ``firm; steady; staunch; resolute''}地\pushnote{声明}{(shēng míng) ``to state; to declare; statement; declaration''}，我从未干过这类事情。}
{I protested my innocence of these things.}
\bothl{
我看得出，在这半是\pushnote{幽默}{(yōu mò) ``(loanword) humor; humorous''}半是\pushnote{挖苦}{(wā kǔ) ``to speak sarcastically; to make cutting remarks''; 挖 is ``to dig'' and 苦 is ``bitter'' as in 辛苦 (xīn kǔ) ``exhausting; hard; tough; arduous; to work hard; to go to a lot of trouble; hardship(s)''}的话语后面，这个老水手，这个与世\pushnote{隔绝}{(gé jué) ``to cut off; to isolate; isolated (from the world)''}的\pushnote{教友}{(jiào yǒu) ``church member''}会\pushnote{信徒}{(xìn tú) ``believer''}式的南塔基人，脑子里装满了岛民的偏见。}{ I saw that under the mask of these half humorous innuendoes, this old seaman, as an insulated Quakerish Nantucketer, was full of his insular prejudices,}
\bothl{除了\pushnote{科德角}{Transliteration: 科 (kē) ``branch of study; administrative section; division; field; branch'' + 德 + 角 (jiǎo) ``angle; corner; horn; horn-shaped''}人和
\pushnote{维因耶德}{Transliteration: 维 (wéi) ``to preserve; to maintain; to hold together; dimension; vitamin'' + 因 (yīn) ``cause; reason; because'' + 耶 (yē) phonetic for ``ye'', seen as part of 椰子 (yē zi) ``coconut'' + 德}人，他们不相信任何外地人。}{ and rather distrustful of all aliens, unless they hailed from Cape Cod or the Vineyard.}

\separator

\bothl{
“说说，你想要去捕鲸和原因是什么呢？

我\pushnote{弄清}{(nòng qīng) ``to clarify; to fully understand''}了这个，才考虑\pushnote{雇}{(gù) ``to employ; to hire; to rent'', commonly used in 雇佣 (gù yōng) ``to employ; to hire''}你当水手。”}{“But what takes thee a-whaling? I want to know that before I think of shipping ye.”}
\bothl{
“好的，先生，我想看看捕鲸是怎么回事？我想去\pushnote{见识}{(jiàn shi) ``to gain first-hand knowledge of sth; to experience for oneself; knowledge; experience; insight''}见识。”}{“Well, sir, I want to see what whaling is. I want to see the world.”}
\bothl{
“想看看捕鲸是怎么回事，是不是，你知道亚哈船长吗？”
}{
“Want to see what whaling is, eh? Have ye clapped eye on Captain Ahab?”
}
\bothl{“亚哈船长是谁，先生？”“}{“Who is Captain Ahab, sir?”}
\bothl{哦，哦，我就知道你不知道。亚哈船长是这船的船长。”}{“Aye, aye, I thought so. Captain Ahab is the Captain of this ship.”}
\bothl{
“这么说，是我弄错了。我还以为我是在和船长本人说话呢。”
}{
“I am mistaken then. I thought I was speaking to the Captain himself.”
}

\bothl{“你是在和\pushnote{法勒}{Transliteration: from 法 (fǎ) ``law; method; way'' + 勒 (lè) ``(literary) bridle; halter; headstall; to rein in; to compel; to force''}船长说话——法勒船长在听你说话，小伙子。”}{“Thou art speaking to Captain Peleg—that's who ye are speaking to, young man. }
\bothl{‘裴廓德号’这次出海，是由我和\pushnote{比勒达}{比 (bǐ) ``to compare'' + 勒 + 达}船长负责准备，备齐船上所需要的一切，包括水手。}{It belongs to me and Captain Bildad to see the Pequod fitted out for the voyage, and supplied with all her needs, including crew. }
\bothl{我们两个都是股东和代理人。}{We are part owners and agents. }
\bothl{不过，我要告诉你，假如你真想知道捕鲸是怎么回事，我\pushnote{倒是}{(dào shi) ``contrary to what one might expect; actually; contrariwise; why don't you''}在你签字\pushnote{画押}{(huà yā) ``to sign; to make one's mark''}\pushnote{绝不反悔}{(jué bù fǎn hǔi) ``to vow not to break one's promise''}之前，有一个\pushnote{法子}{(fǎ zi) ``way; method''}让你了解一下。}{But as I was going to say, if thou wantest to know what whaling is, as thou tellest ye do, I can put ye in a way of finding it out before ye bind yourself to it, past backing out.”}

\separator

\bothl{小伙子，去看看亚哈船长吧，你会发现他只有一条腿。}{Clap eye on Captain Ahab, young man, and thou wilt find that he has only one leg.”}
\bothl{
“你的意思是……被大鲸给弄去了一条？”“
}{
“What do you mean, sir? Was the other one lost by a whale?”
}
\bothl{
是大鲸弄去的！小伙子，我跟你说：

是一条曾经把一只小艇打得粉碎的无比

\pushnote{凶恶}{(xiōng è) ``fierce; ferocious; fiendish; frightening''}的\pushnote{抹香鲸}{(mǒ xiāng jīng) ``sperm whale; cachalot (Physeter macrocephalus)''}把腿\pushnote{咬}{(yǎo) ``to bite; to nip''}断，\pushnote{嚼}{(jiáo) ``to chew''}\pushnote{碎}{(suì) ``(transitive or intransitive verb) to break into pieces; to shatter; to crumble; broken; fragmentary; scattered''}，嚼得\pushnote{咯吱}{(gē zhī) ``(onom.) creak; groan''}咯吱\pushnote{响}{(xiǎng) ``echo; sound; noise; to make a sound; to sound; to ring; loud; classifier for noises''}！——\pushnote{唉}{(āi) ``interjection or grunt of agreement or recognition (e.g. yes, it's me!); to sigh''}！唉！”}{“Lost by a whale! Young man, come nearer to me: it was devoured, chewed up, crunched by the monstrousest parmacetty that ever chipped a boat!—ah, ah!”
}

\separator


\bothl{他\pushnote{憋足了劲}{(biē zú le jìn) idiom: ``reach the limit, use up all the energy''}的\pushnote{神态}{(shén tài) ``appearance; manner; bearing; deportment; look; expression''}\pushnote{着实}{(zhuó shí) ``truly; indeed; severely; harshly''}把我\pushnote{镇住}{(zhèn zhù) ``to dominate; to control; to subdue; to crush''}了，最后两声\pushnote{发自}{(fā zì) ``to evolve from''}内心的\pushnote{痛苦}{(tòng kǔ) ``pain; suffering; painful''}\pushnote{呻吟}{(shēn yín) ``to moan; to groan''}，也很让我为之\pushnote{震撼}{(zhèn hàn) ``to shake; to shock; to stun; shocking; stunning; shock''}，但我尽量保持神态自然，说：}{I was a little alarmed by his energy, perhaps also a little touched at the hearty grief in his concluding exclamation, but said as calmly as I could, }
\bothl{“你说的肯定错不了，先生，可是我怎么知道那条鲸竟会如此\pushnote{十恶不赦}{(shí è bù shè) ``wicked beyond redemption (idiom); heinous''}呢，说实在的，我可以从这简单的事故中\pushnote{推知}{(tūi zhī) ``to infer''}许多\pushnote{情形}{(qíng xing) ``circumstances; situation''}。”}{“What you say is no doubt true enough, sir; but how could I know there was any peculiar ferocity in that particular whale, though indeed I might have inferred as much from the simple fact of the accident.”}

\separator

\bothl{“你看你，小伙子，你还\pushnote{嫩}{(nèn) ``young and tender; (of food) tender; lightly cooked; (of color) light; (of a person) inexperienced; unskilled''}了点，明白吗；不过还数你不好\pushnote{吹嘘}{(chuī xū) ``to brag''}。当然啰，你是出过海的，可这算得了什么？”}{“Look ye now, young man, thy lungs are a sort of soft, d'ye see; thou dost not talk shark a bit. Sure, ye've been to sea before now; sure of that?”}
\bothl{“先生，”我说，“我记得我跟你说出过四次海，是在商——”}{“Sir,” said I, “I thought I told you that I had been four voyages in the merchant—”}
\bothl{“你别再说了！记得对商船水手有何感想吗——别惹我烦——我不想听。}{“Hard down out of that! Mind what I said about the marchant service—don't aggravate me—I won't have it. }
\bothl{不过，话总得说明白。我已经给你\pushnote{挑明}{(tiǎo míng) ``to illuminate; to open up (a topic)''}了捕鲸是怎么回事？你还想去捕鲸吗？”}{But let us understand each other. I have given thee a hint about what whaling is; do ye yet feel inclined for it?”}
\bothl{“没错，先生。”}{“I do, sir.”}
\bothl{“很好，那么，你敢把鱼\pushnote{叉}{(chā) ``fork; pitchfork; prong; pick; cross; intersect''}\pushnote{刺}{(cì) ``thorn; sting; thrust; to prick; to pierce; to stab; to assassinate; to murder''}进一条活鲸的\pushnote{喉咙}{(hóu lóng) ``throat''}，再\pushnote{穷追不舍}{(qióng zhuī bù shě) ``to pursue relentlessly''}
吗？你说，快告诉我！”}{“Very good. Now, art thou the man to pitch a harpoon down a live whale's throat, and then jump after it? Answer, quick!”}
\bothl{“我敢，先生。如果非这么干不可的话，或者说不这样就会被鲸干掉的话，我会敢的。可我认为不会有这种情况。”}{“I am, sir, if it should be positively indispensable to do so; not to be got rid of, that is; which I don't take to be the fact.”}

\end{paracol}

\outputnotes

\end{document}
